---
title: "SPIAT worksheet"
author: "Yuzhou Feng"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome

Welcome to the SPIAT spatial analysis workshop! In this worksheet we
will explore how to analyse spatial proteomics data using SPIAT and simulate 
spatial point pattern using spaSim. Let's get started!

This worksheet was adapted from the [SPIAT tutorials](https://trigosteam.github.io/SPIAT/articles/SPIAT.html), but
uses a real dataset instead. Please refer to the tutorials for
detailed information.

Please make sure you have installed all the required packages prior to
the workshop. We won't have time for installation during the workshop.

```{r install, message=FALSE}
## Install the packages: 
## (Only run the code if the packages were not installed yet)
## (latest verion of R required >=4.3.0)

# install.packages(c("elsa", "Rtsne", "umap", "alphahull", "plotly", "survminer", "survival", "vroom"))
# install.packages(c("rmarkdown", "knitr"))
# 
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install(version = "3.18")
# 
# BiocManager::install("SPIAT")
# BiocManager::install("spaSim")
# BiocManager::install("ComplexHeatmap")
```

```{r load, message=FALSE}
# load the libraries
library(SPIAT)
library(spaSim)
```

In this workshop we will demonstrate the analysis workflow with a public
TNBC dataset (Keren et al., 2018).

# Reading and formatting

## Reading and formatting

`format_image_to_spe()` is the main function to read in data to SPIAT.
`format_image_to_spe()` creates a `SpatialExperiment` object which is
used in all subsequent functions. The key data points of interest for
SPIAT are cell coordinates, marker intensities and cell phenotypes for
each cell.

`format_image_to_spe()` has several options designed specifically for
importing data generated from the inForm, HALO, CODEX and cellprofiler
platforms. However, we advise using the more flexible 'general' option.
Note that before using 'general', we advise pre-formatting your data.

```{r}
# use `?` to get the usage of a function in the "help" panel
?format_image_to_spe
```

Now we'll use the `general` option to read in the TNBC MIBI data. This
MIBI file is not the raw MIBI data, but was generated after cell
segmentation (by Greg Bass, CSL). This shows SPIAT's capability of
reading data from a wide range of formats.

```{r}
# read the csv
path_to_file <- "cellData_with-centroids.csv" # put the path to the csv in the quote
mibi <- vroom::vroom(path_to_file)
mibi[1:5,] # look at the first 5 rows
unique(mibi$SampleID) # how many unique samples here?
```

```{r}
# slice the data frame to keep only the sample with `sampleID==4`
mibi_4 <- mibi[mibi$SampleID == 4, ]
dim(mibi_4)
# add a column of cell IDs
mibi_4$CellID <- paste("Cell", mibi_4$cellLabelInImage, sep="_")
```

```{r}
# construct the elements of the SPE object
## the marker intensity matrix
intensity <- t(mibi_4[,13:52])
intensity <- rbind(intensity, Tumor = mibi_4$tumorYN)
intensity <- intensity[rownames(intensity) %in% c("CD4", "CD3", "CD8",
                                                "CD20", "MPO", "CD68",
                                                "Tumor"),]
colnames(intensity) <- mibi_4$CellID # add cell IDs
intensity[, 1:5] # print the intensity of first 5 cells
## the coordinates
coord_x <- mibi_4$Xc
coord_y <- mibi_4$Yc
## we don't have phenotype info in the file. Will predict the phenotypes later

# now format the image with function `format_image_to_spe()`
mibi_4_spe <- format_image_to_spe(format = "general", phenotypes = NA,
                                  intensity_matrix = intensity,
                                  coord_x = coord_x, coord_y = coord_y)
```

Now we get the SPE object of the image with `sampleID==4`.

## Predicting phenotypes

You'll notice there is no phenotype info in the MIBI .csv file. Let's
use SPIAT to predict phenotypes based on the marker intensity
distribution of all cells.

```{r}
# Print the markers to predict
rownames(assay(mibi_4_spe))
# Predict (`reference_phenotypes = FALSE` because there are no default phenotypes)
mibi_4_spe <- predict_phenotypes(mibi_4_spe, tumour_marker = "Tumor",
                   baseline_markers = c("CD4", "CD68", "CD8", "CD3", "CD20", "MPO"),
                   reference_phenotypes = FALSE)
```

Have a look at the structure of this object. There are three main
measurements of interest (coordinates, phenotype, intensities).

```{r}
# View(mibi_4_spe)
# Access phenotype as a column (accessed with $)
# Print the unique phenotypes
unique(mibi_4_spe$Phenotype)
# What is the dimension of the intensity matrix of this object?
dim(assay(mibi_4_spe))
# Access marker intensities with the helper function 'assay'
# Print the marker intensities of the first 5 cells
assay(mibi_4_spe)[, 1:5]
# Access coordinates with the helper function 'spatialCoords'
spatialCoords(mibi_4_spe)[1:5, ]
```

## Define cell types

Then we define the cell types.

```{r}
phenotypes <- c("Tumor", "CD68", "CD20", "CD4,CD3", "CD3", "CD8,CD3", 
                "CD4", "MPO", "CD8")
cell_types <- c("Tumor", "Macrophage", "B_cells",  "Helper_T_cell", "T_cell",
                "Cyto_T_cell",  "Helper_T_cell", "Neutrophil", "Cyto_T_cell")

defined_mibi_4 <- define_celltypes(
    mibi_4_spe, 
    categories = phenotypes, 
    category_colname = "Phenotype", 
    names = cell_types,
    new_colname = "Cell.Type")
unique(defined_mibi_4$Cell.Type)
```

You can write a `for` loop to read all of the patients' images. We will
look into that later.

# Quality control and data visualisation

We'll now use SPIAT's quality control functions to check phenotyping
quality, detect uneven staining, and test for other potential technical
artefacts.

## Visualise marker levels

Phenotyping of cells can be verified comparing marker intensities of
cells labelled positive and negative for a marker. Cells positive for a
marker should have high levels of the marker. An unclear separation of
marker intensities between positive and negative cells would suggest
phenotypes have not been accurately assigned. We can use
`marker_intensity_boxplot()` to produce a boxplot for cells phenotyped
as being positive or negative for a marker.

```{r}
markers <- rownames(assay(mibi_4_spe))
for (marker in markers){
  print(marker_intensity_boxplot(mibi_4_spe, marker))
}
```

Uneven marker staining or high background intensity can be identified
with `plot_cell_marker_levels()`. This produces a scatter plot of the
intensity of a marker in each cell. This should be relatively even
across the image and all phenotyped cells. Cells that were not
phenotyped as being positive for the particular marker are excluded.

```{r}
plot_cell_marker_levels(mibi_4_spe, "CD3")
# Try plot all markers with a for loop~
for (marker in markers){
  print(plot_cell_marker_levels(mibi_4_spe, marker))
}
```

For large images, there is also the option of 'blurring' the image with
function `plot_marker_level_heatmap()`, where the image is split into
multiple small areas, and marker intensities are averaged within each.
The image is blurred based on the `num_splits` parameter.

```{r}
# Try it yourself~
```

## Identifying incorrect phenotypes

We may see cells with biologically implausible combination of markers
present in the input data when using:

```{r}
unique(mibi_4_spe$Phenotype)
```

Removing cells with incorrect phenotypes: If you identify incorrect
phenotypes or have any properties (columns) that you want to exclude you
can use `select_celltypes()`. Set `celltypes` as the values that you
want to keep or exclude for a column (`feature_colname`). Set `keep` as
`TRUE` to include these cells and `FALSE` to exclude them.

```{r}
data_subset <- select_celltypes(
  mibi_4_spe, keep=TRUE,
  celltypes = c("Tumor", "CD68", "CD20", "CD4,CD3", 
                 "CD3", "CD8,CD3", "CD4", "MPO", "CD8"),
  feature_colname = "Phenotype")
# have a look at what phenotypes are present
unique(data_subset$Phenotype)
# All the cells with the rest of the combinations of phenotypes were removed.
# define cell types (try it yourself~)
defined_mibi_subset <- define_celltypes(mibi_4_spe, category_colname = "Phenotype"
)
```

## Dimensionality reduction to identify misclassified cells

We can also check for specific misclassified cells using dimensionality
reduction. SPIAT offers tSNE and UMAPs based on marker intensities to
visualise cells. Cells of distinct types should form clearly different
clusters.

The generated dimensionality reduction plots are interactive, and users
can hover over each cell to obtain the cell ID. Users can then remove
specific misclassified cells.

```{r}
g <- dimensionality_reduction_plot(defined_mibi_subset, plot_type = "TSNE", 
                                   feature_colname = "Cell.Type")
```

Note that `dimensionality_reduction_plot()` only prints a static version
of the UMAP or tSNE plot. To interact with this plot, pass the result to
the `ggplotly()` function from the `plotly` package.

```{r}
plotly::ggplotly(g)
```

## Visualising tissues

Let's go back to the image without filtering out the undefined cells -
`mibi_4_spe`.

We can see the location of all cell types (or any column in the data) in
the tissue with `plot_cell_categories()`. Each dot in the plot
corresponds to a cell and cells are coloured by cell type. Any cell
types present in the data but not in the cell types of interest will be
put in the category "OTHER" and coloured lightgrey.

```{r}
my_colors <- c("red", "blue","orange", "purple", "pink", "darkcyan", "darkgreen")
categories_of_interest <- c("Tumor", "Macrophage", "B_cells",  "Helper_T_cell", 
                            "T_cell", "Cyto_T_cell", "Neutrophil")
  
plot_cell_categories(spe_object = defined_mibi_4, 
                     categories_of_interest = categories_of_interest, 
                     colour_vector = my_colors, feature_colname = "Cell.Type")
```

We can visualise a selected marker in 3D with `marker_surface_plot()`.
The image is blurred based on the `num_splits` parameter.

```{r}
marker_surface_plot(defined_mibi_4, num_splits=15, marker="CD68")
```

To visualise multiple markers in 3D in a single plot we can use
`marker_surface_plot_stack()`. This shows normalised intensity level of
specified markers and enables the identification of co-occurring and
mutually exclusive markers.

```{r}
marker_surface_plot_stack(defined_mibi_4, num_splits=15, markers_to_plot=c("CD68", "CD20"))
```

# Basic analyses

## Cell percentages

Obtain the number and proportion of each cell type. We can exclude any
cell types that are not of interest e.g. "Undefined" with
`celltypes_to_exclude`.

```{r}
p_cells <- calculate_cell_proportions(defined_mibi_4, 
                                      feature_colname ="Cell.Type",
                                      celltypes_to_exclude = c("Others","Undefined"),
                                      plot.image = TRUE)
```

```{r}
plot_cell_percentages(cell_proportions = p_cells, 
                      cells_to_exclude = "Tumor", cellprop_colname="Proportion_name")
```

## Cell distances

We can calculate the pairwise distances between two cell types (cell
type A and cell type B) with
`calculate_pairwise_distances_between_cell_types()`. This function
calculates the distances of all cells of type A against all cells of
type B.

```{r}
distances <- calculate_pairwise_distances_between_celltypes(
  spe_object = defined_mibi_4, 
  cell_types_of_interest = c("Tumor", "B_cells", "T_cell"),
  feature_colname = "Cell.Type")
```

```{r}
# Visualise the distances
plot_cell_distances_violin(distances)
```

We can also calculate summary statistics for the distances between each
combination of cell types, the mean, median, min, max and standard
deviation, with `calculate_summary_distances_between_celltypes()`.

```{r}
summary_distances <- calculate_summary_distances_between_celltypes(distances)
summary_distances
```

An example of the interpretation of this result is: "average pairwise
distance between cells of Tumor and B cells is 1107.9081.

These pairwise cell distances can then be visualised as a heatmap with
plot_distance_heatmap(). This example shows the average pairwise
distances between cell types. Note that the pairwise distances are
symmetrical (the average distance between cell type A and cell type B is
the same as the average distance between cell Type B and cell Type A).

```{r}
plot_distance_heatmap(phenotype_distances_result = summary_distances, metric = "mean")
```

The same workflow also applies to minimum distance between cell types.
Give it a go with the following functions:
`calculate_minimum_distances_between_celltypes()`,
`plot_cell_distances_violin()`,
`calculate_summary_distances_between_celltypes()`,
`plot_distance_heatmap()`.

```{r}
# Try these yourself~
min_distances = calculate_minimum_distances_between_celltypes(spe_object = defined_mibi_4, 
  cell_types_of_interest = c("Tumor", "B_cells", "T_cell"),
  feature_colname = "Cell.Type")
plot_cell_distances_violin(min_distances)
min_distances_sum =calculate_summary_distances_between_celltypes(min_distances)
plot_distance_heatmap(min_distances_sum, metric = "mean")
```

# Cell colocalisation

With SPIAT we can quantify cell colocalisation, which refers to how much
two cell types are colocalising and thus potentially interacting.

## Cells in Neighbourhood (CIN)

We can calculate the average percentage of cells of one cell type
(target) within a radius of another cell type (reference) using
`average_percentage_of_cells_within_radius()`.

```{r}
average_percentage_of_cells_within_radius(spe_object = defined_mibi_4, 
                                          reference_celltype = "B_cells", 
                                          target_celltype = "Macrophage", 
                                          radius=100, feature_colname="Cell.Type")
```

Alternatively, this analysis can also be performed based on marker
intensities rather than cell types. Here, we use
`average_marker_intensity_within_radius()` to calculate the average
intensity of the target_marker within a radius from the cells positive
for the reference marker. Note that it pools all cells with the target
marker that are within the specific radius of any reference cell.
Results represent the average intensities within a radius.

```{r}
average_marker_intensity_within_radius(spe_object = defined_mibi_4,
                                       reference_marker ="CD20",
                                       target_marker = "CD68",
                                       radius=100)
```

`plot_average_intensity()` calculates the average intensity of a target
marker for a number of user-supplied radii values, and plots the
intensity level at each specified radius as a line graph.

```{r}
plot_average_intensity(spe_object=defined_mibi_4, 
                       reference_marker="CD20", 
                       target_marker="CD68", radii=seq(20, 100, 5))
```

What can you interpret from this plot?

## Mixing Score (MS) and Normalised Mixing Score (NMS)

Mixing score (MS) in SPIAT is defined as the number of target-reference
interactions divided by the number of reference-reference interactions
within a specified radius. The higher the score the greater the mixing
of the two cell types. The normalised score is normalised for the number
of target and reference cells in the image.

```{r}
# What is the mixing score between Tumor cells and Macrophage in our sample?
mixing_score_summary(spe_object = defined_mibi_4, 
                     reference_celltype = "Tumor", 
                     target_celltype = "Macrophage", 
                     radius=20, feature_colname ="Cell.Type")
```

## Cross K function

Cross K function calculates the number of target cell types across a
range of radii from a reference cell type, and compares the behaviour of
the input image with an image of randomly distributed points using a
Poisson point process. There are four patterns that can be distinguished
from K-cross function - independent, mixed clusters, mixed one cluster,
separate clusters.

Try different cell types with different radii using
`calculate_cross_functions()` . What patterns do you identify?

```{r}
df_cross <- calculate_cross_functions(defined_mibi_4, method = "Kcross", 
                                      cell_types_of_interest = c("Tumor", "B_cells"), 
                                      feature_colname ="Cell.Type",
                                      dist = 500)
```

We can calculate the area under the curve (AUC) of the cross K-function.
In general, this tells us the two types of cells are:

-   negative values: separate clusters

-   positive values: mixing of cell types

```{r}
AUC_of_cross_function(df_cross)
```

## Cross-K Intersection (CKI)

There is another pattern in cross K curve which has not been previously
appreciated, which is when there is a "ring" of one cell type, generally
immune cells, surrounding the area of another cell type, generally
tumour cells. For this pattern, the observed and expected curves in
cross K function cross or intersect, such as the cross K plot above.

While this pattern does not appear in the example data, you can
experiment with this pattern using the simulated data from the SPIAT
package, or refer to the [CKI section in the
tutorial](https://trigosteam.github.io/SPIAT/articles/cell-colocalisation.html#cross-k-intersection-cki).

## Aggregated entropy gradient

The function `entropy_gradient_aggregated()` proposes the aggregated
entropy gradient as a self-contained metric to define the attraction and
repulsion of cell types. You don't have to set arbitrary threshold, but
only use the shape of the curve to determine the colocalisation. The
details of this algorithm were discussed thoroughly in Fig. 4 and
Supplementary Fig. 4 of the
[paper](https://www.nature.com/articles/s41467-023-37822-0).

Let's have a look what results this function gives us.

```{r}
table(defined_mibi_4$Cell.Type)
gradient_pos <- c(50, 75, 100, 125, 150, 175, 200, 250, 300, 350, 400, 450, 500, 550, 600)
grad <- entropy_gradient_aggregated(defined_mibi_4, 
                                    cell_types_of_interest = c("Macrophage", "Tumor"),
                                    feature_colname = "Cell.Type",
                                    radii =  gradient_pos)

grad
```

Plot the gradients.

```{r}
v <- as.numeric(grad$gradient_df[1,3:(length(gradient_pos)+2)])
plot(v, type = "b", lty = 2, pch = 16, cex = 1)
```

The trend indicates there is repulsion between tumor and macrophages.
Let's re-visit the cell category plot of the image and see the locations
of these cells.

```{r}
plot_cell_categories(spe_object = defined_mibi_4, 
                     categories_of_interest = categories_of_interest, 
                     colour_vector = my_colors, feature_colname = "Cell.Type")
```

Note that the shape of the curve is determined by several factors,
including using which cell type as the reference cells. The discussion
of this is included in the [Supplementary Fig. 4](https://www.nature.com/articles/s41467-023-37822-0/figures/4) of the paper.

# Spatial Heterogeneity

Cell colocalisation metrics allow capturing a dominant spatial pattern
in an image. However, patterns are unlikely to be distributed evenly in
a tissue, but rather there will be spatial heterogeneity of patterns. To
measure this, SPIAT splits the image into smaller images (either using a
grid or concentric circles around a reference cell population), followed
by calculation of a spatial metric of a pattern of interest (e.g. cell
colocalisation, entropy), and then measures the Prevalence and
Distinctiveness of the pattern.

## Entropy

Entropy in spatial analysis refers to the balance in the number of cells
of distinct populations. An entropy score can be obtained for an entire
image. However, the entropy of one image does not provide us spatial
information of the image.

```{r}
calculate_entropy(defined_mibi_4, cell_types_of_interest = c("Tumor","Macrophage"), 
                  feature_colname = "Cell.Type")
```

We therefore propose the concept of Localised Entropy which calculates
an entropy score for a predefined local region.

## Fishnet grid

One approach to calculate localised metric is to split the image into
fishnet grid squares. For each grid square, `grid_metrics()` calculates
the metric for that square and visualise the raster image. You can
choose any metric as the localised metric. Here we use entropy as an
example.

For cases where the localised metric is not symmetrical (requires
specifying a target and reference cell type), the first item in the
vector used for `cell_types_of_interest` marks the reference cells and
the second item the target cells. Here we are using Entropy, which is
symmetrical, so we can use any order of cell types in the input.

```{r}
grid <- grid_metrics(defined_mibi_4, FUN = calculate_entropy, n_split = 20,
                     cell_types_of_interest=c("Tumor","Macrophage"), 
                     feature_colname = "Cell.Type")
```

After calculating the localised entropy for each of the grid squares, we
can apply metrics like percentages of grid squares with patterns
(Prevalence) and Moran's I (Distinctiveness).

For the Prevalence, we need to select a threshold over which grid
squares are considered 'positive' for the pattern. The selection of
threshold depends on the pattern and metric the user chooses to find the
localised pattern. Here we chose 0.73 for entropy because 0.73 is
roughly the entropy of two cell types when their ratio is 1:4 or 4:1.

You might be curious of how the entropy is calculated based on the
ratio. Use the following helper function to calculate the entropy for
any composition of cell counts.

```{r entropy_helper}
entropy_helper <- function(v){
   s <- sum(v); l <- length(v); e <- 0
   for (i in 1:l) e <- e + -(v[i]/s)*log2(v[i]/s) 
   return(e)
 }
```

What is the entropy when the ratio of two cell types is 1:4?

```{r}
entropy_helper(c(1,4))
entropy_helper(c(1,8))
entropy_helper(c(1,16))
```

```{r}
calculate_percentage_of_grids(grid, threshold = 0.73, above = TRUE)
calculate_spatial_autocorrelation(grid, metric = "globalmoran")
```

# Characterise tissue structure

In certain analysis the focus is to understand the spatial distribution
of a certain type of cell populations relative to the tissue regions.

One example of this functionality is to characterise the immune
population in tumour structures. The following analysis will focus on
the tumour/immune example, including determining whether there is a
clear tumour margin, automatically identifying the tumour margin, and
finally quantifying the proportion of immune populations relative to the
margin. However, these analyses can also be generalised to other tissue
and cell types.

### Determining whether there is a clear tumour margin

In some instances tumour cells are distributed in such a way that there
are no clear tumour margins. While this can be derived intuitively in
most cases, SPIAT offers a way of quantifying the 'quality' of the
margin for downstream analyses. This is meant to be used to help flag
images with relatively poor margins, and therefore we do not offer a
cutoff value.

To determine if there is a clear tumour margin, SPIAT can calculate the
ratio of tumour bordering cells to tumour total cells (R-BC). This ratio
is high when there is a disproportional high number of tumour margin
cells compared to internal tumour cells.

```{r}
R_BC(defined_mibi_4, cell_type_of_interest = "Tumor", "Cell.Type")
```

The result is 0.1508639. This low value means there are relatively low
number of bordering cells compared to total tumour cells, meaning that
this image has clear tumour margins.

### Automatic identification of the tumour margin

We can identify margins with `identify_bordering_cells()`. This function
leverages off the alpha hull method (Pateiro-Lopez, Rodriguez-Casal,
and. 2019) from the alpha hull package. Here we use tumour cells
(Tumour_marker) as the reference to identify the bordering cells but any
cell type can be used.

```{r}
formatted_border <- identify_bordering_cells(defined_mibi_4, 
                                             reference_cell = "Tumor", 
                                             feature_colname="Cell.Type",
                                             ahull_alpha = 30,
                                             n_to_exclude = 25)
# Try with different ahull_alpha values and n_to_exclude values

# Then get the number of tumour clusters
attr(formatted_border, "n_of_clusters")
```

There are 3 tumour clusters in the image.

### Classification of cells based on their locations relative to the margin

First, we calculate the distance of cells to the tumour margin.

```{r}
formatted_distance <- calculate_distance_to_margin(formatted_border)
```

Next, we classify cells based on their location. As a distance cutoff,
we use a distance of 5 cells from the tumour margin. The function first
calculates the average minimum distance between all pairs of nearest
cells and then multiples this number by 5. Users can change the number
of cell layers to increase/decrease the margin width or use another
argument `margin_dist`.

```{r}
names_of_immune_cells <- c("B_cells",  "Helper_T_cell", "T_cell",
                "Cyto_T_cell",  "Helper_T_cell", "Neutrophil", "Cyto_T_cell")

formatted_structure <- define_structure(
  formatted_distance, cell_types_of_interest = names_of_immune_cells, 
  feature_colname = "Cell.Type", margin_dist = 90)

categories <- unique(formatted_structure$Structure)
```

We can plot and colour these structure categories.

```{r}
plot_cell_categories(formatted_structure, feature_colname = "Structure")
```

Then calculate the proportions of immune cells in each of the locations.

```{r}
immune_proportions <- calculate_proportions_of_cells_in_structure(
  spe_object = formatted_structure, 
  cell_types_of_interest = names_of_immune_cells, feature_colname ="Cell.Type")

immune_proportions
```

Finally calculate summaries of the distances for immune cells in the
tumour structure.

```{r}
immune_distances <- calculate_summary_distances_of_cells_to_borders(
  spe_object = formatted_structure, 
  cell_types_of_interest = names_of_immune_cells, feature_colname = "Cell.Type")

immune_distances
```

# Cellular neighbourhood

The aggregation of cells can result in cellular neighbourhoods. A
neighbourhood is defined as a group of cells that cluster together.
These can be homotypic, containing cells of a single class (e.g.immune
cells), or heterotypic (e.g.a mixture of tumour and immune cells).

Function `identify_neighborhoods()` identifies cellular neighbourhoods.
Users can select a subset of cell types of interest if desired. The
algorithm is *Hierarchical Clustering algorithm* - Euclidean distances
between cells are calculated, and pairs of cells with a distance less
than a specified radius are considered to be interacting, with the rest
being non-interacting. Hierarchical clustering is then used to separate
the clusters. Larger radii will result in the merging of individual
clusters.

You need to specify a radius that defines the distance for an
interaction. We suggest testing different radii and select the one that
generates intuitive clusters upon visualisation. Cells not assigned to
clusters are assigned as `Cluster_NA` in the output table. The argument
`min_neighborhood_size` specifies the threshold of a neighborhood size
to be considered as a neighborhood. Smaller neighbourhoods will be
outputted, but will not be assigned a number.

Please test out different radii and then visualise the clustering
results. To aid in this process, users can use the
`average_minimum_distance()` function, which calculates the average
minimum distance between all cells in an image, and can be used as a
starting point.

```{r}
average_minimum_distance(defined_mibi_4)
```

We then identify the cellular neighbourhoods using our hierarchical
algorithm with a radius of 50, and with a minimum neighbourhood size of
100. Cells assigned to neighbourhoods smaller than 100 will be assigned
to the Cluster_NA neighbourhood.

```{r}
clusters <- identify_neighborhoods(
  defined_mibi_4, method = "hierarchical", min_neighborhood_size = 100,
  cell_types_of_interest = c("B_cells",  "Helper_T_cell", "T_cell",
                "Cyto_T_cell",  "Helper_T_cell", "Neutrophil", "Cyto_T_cell"), radius = 50, 
  feature_colname = "Cell.Type")
```

Black cells correspond to 'free', un-clustered cells.

We can visualise the cell composition of neighborhoods. To do this, we
can use `composition_of_neighborhoods()` to obtain the percentages of
cells with a specific marker within each neighborhood and the number of
cells in the neighborhood.

In this example we select cellular neighbourhoods with at least 5 cells.

```{r}
neighorhoods_vis <- 
  composition_of_neighborhoods(clusters, feature_colname = "Cell.Type")
neighorhoods_vis <- 
  neighorhoods_vis[neighorhoods_vis$Total_number_of_cells >=5,]
```

Finally, we can use `plot_composition_heatmap()` to produce a heatmap
showing the marker percentages within each cluster, which can be used to
classify the derived neighbourhoods.

```{r}
plot_composition_heatmap(neighorhoods_vis, feature_colname="Cell.Type")
```

# Challenge

So far we have walked through the analysis workflow. Now it's your turn
to do some analyses!

The TNBC MIBI data set provides the clinical information of 39 patients.
Try applying some spatial metrics to these images and associate them
with the clinical data!

The SPE objects have been pre-formatted for you. Please load the objects:

```{r}
load("MIBI_spe.Rda") # load the MIBI_spe objects
# The code of formatting the images is in SPIAT_worksheet_solutions.Rmd file
```

Here is only an example of what you can do with the SPIAT functions. Please
follow me to complete the code.

```{r, message = FALSE}
# calculate the mixing score of each patient
mixing_score <- list()
normalised_mixing_score <- list()
for(patient in names(MIBI_spe)){
   ms <- mixing_score_summary(
    spe_object = MIBI_spe[[patient]], reference_celltype = "Tumor",
    target_celltype = "Macrophage", radius = 20, feature_colname = "Cell.Type2")
  mixing_score[[patient]] <- ms$Mixing_score
  normalised_mixing_score[[patient]] <- ms$Normalised_mixing_score
  print(patient)
}
# group the patients based on their normalised mixing score
nms <- na.omit(unlist(normalised_mixing_score))

low_nms <- normalised_mixing_score < 0.25 # this is an arbitrary threshold, adopted from the spaSim benchmarking
group_repulsion <- names(which(low_nms == TRUE))
group_attraction <- names(which(low_nms == FALSE))

# read in metadata
metadata <- read.delim("Formatted_clinical_data.csv", sep = ",")
metadata$Patient <- paste("Patient", metadata$InternalId, sep="_")
rownames(metadata) <- metadata$Patient

# join the data
metadata$Mixing_group <- NA
metadata$Mixing_group[rownames(metadata) %in% group_repulsion] <- "Group_Repulsion"
metadata$Mixing_group[rownames(metadata) %in% group_attraction] <- "Group_Attraction"

# survival
library(survminer)
library(survival)
survival <- survfit(Surv(Survival_days_capped., Censored) ~ Mixing_group, data=metadata)
  
g <- ggsurvplot(survival, 
             xlab = "Days", 
             ylab = "Overall survival probability", pval=TRUE)
print(g)
```

# Introduction to spaSim

`spaSim` (**spa**tial **Sim**ulator) is a simulator of tumour immune
microenvironment spatial data. It includes a family of functions to
simulate a diverse set of cell localisation patterns in tissues.
Patterns include background cells (one cell type or multiple cell types
of different proportions), tumour/immune clusters, immune rings and
double immune rings and stripes (blood/lymphatic vessels).

Simulations from spaSim can be applied to test and benchmark spatial
tools and metrics (like the ones in `SPIAT`). The output of spaSim are
images in `SpatialExperiment` object format.

In this workshop, we will only cover how to simulate individual images.
For more information about how to simulate a set of images, please refer to the 
[spaSim tutorial](https://trigosteam.github.io/spaSim/articles/vignette.html).

```{r}
# simulate background cells
set.seed(610)
bg <- simulate_background_cells(n_cells = 5000,
                                width = 2000,
                                height = 2000,
                                method = "Hardcore",
                                min_d = 10,
                                oversampling_rate = 1.6,
                                Cell.Type = "Others")
```

```{r}
# simulate mixed background
mix_bg <- simulate_mixing(bg_sample = bg,
                          idents = c("Tumour", "Immune", "Others"),
                          props = c(0.2, 0.3, 0.5), 
                          plot_image = TRUE,
                          plot_colours = c("red","darkgreen","lightgray"))
```

```{r}
# simulate clusters
cluster_properties <- list(
  C1 =list(name_of_cluster_cell = "Tumour", size = 500, shape = "Oval", 
           centre_loc = data.frame(x = 600, y = 600),infiltration_types = c("Immune1", "Others"), 
           infiltration_proportions = c(0.1, 0.05)), 
  C2 = list(name_of_cluster_cell = "Immune1", size = 600,  shape = "Irregular", 
            centre_loc = data.frame(x = 1500, y = 500), infiltration_types = c("Immune", "Others"),
            infiltration_proportions = c(0.1, 0.05)))
# can use any defined image as background image, here we use mix_bg defined in the previous section
clusters <- simulate_clusters(bg_sample = mix_bg,
                              n_clusters = 2,
                              bg_type = "Others",
                              cluster_properties = cluster_properties,
                              plot_image = TRUE,
                              plot_categories = c("Tumour" , "Immune", "Immune1", "Others"),
                              plot_colours = c("red", "darkgreen", "darkblue", "lightgray"))
```

```{r}
# simulate immune rings
immune_ring_properties <- list(
  I1 = list(name_of_cluster_cell = "Tumour", size = 500, 
            shape = "Circle", centre_loc = data.frame(x = 930, y = 1000), 
            infiltration_types = c("Immune1", "Immune2", "Others"), 
            infiltration_proportions = c(0.15, 0.05, 0.05),
            name_of_ring_cell = "Immune1", immune_ring_width = 150,
            immune_ring_infiltration_types = c("Immune2", "Others"), 
            immune_ring_infiltration_proportions = c(0.1, 0.15)))
rings <- simulate_immune_rings(
  bg_sample = bg,
  bg_type = "Others",
  n_ir = 1,
  ir_properties = immune_ring_properties,
  plot_image = TRUE,
  plot_categories = c("Tumour", "Immune1", "Immune2", "Others"),
  plot_colours = c("red", "darkgreen", "darkblue", "lightgray"))
```

